---
title: "Reproducible Research: Peer Assessment 1"
author: Mike B
date: "2025-11-03"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading packages all at one time. Added extras just in case.   

```{r load-packages, message=FALSE, warning=FALSE}
# Load all packages at one time (added extras just in case)
pkgs <- c("ggplot2","dplyr","tidyr","readr","purrr","tibble","stringr","forcats",
          "lubridate","knitr","rmarkdown","gridExtra","scales")
install_if_missing <- function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) 
  install.packages(pkg, repos = "https://cloud.r-project.org")
invisible(lapply(pkgs, install_if_missing))
invisible(lapply(pkgs, library, character.only = TRUE))
```

## Loading and Preprocessing the data.
```{r load-data}
unzip("activity.zip")

activity <- readr::read_csv(
  "activity.csv",
  col_types = readr::cols(
    steps    = readr::col_double(),
    date     = readr::col_date(format = "%Y-%m-%d"),
    interval = readr::col_integer()
  )
)

# Quick glance at structure
glimpse(activity)

# Count total missing values
sum(is.na(activity$steps))
```

## What is mean total number of steps taken per day?

## Calculates "Total Daily Steps" while ignoring missing values.
## Plots a simple, clean Histogram.

```{r mean-steps-per-day}
# Calculate total steps per day -------------------------------------------
daily_steps <- activity %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))

# Histogram of total daily steps ------------------------------------------
ggplot(daily_steps, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "#2c7bb6", color = "white") +
  labs(title = "Total Steps per Day",
       x = "Steps per Day",
       y = "Frequency")

# Mean and median total steps per day -------------------------------------
mean_steps <- mean(daily_steps$total_steps)
median_steps <- median(daily_steps$total_steps)

mean_steps
median_steps

```

## What is the average daily activity pattern?
```{r avg-daily-pattern}
# Calculate the average number of steps for each 5-minute interval ---------
avg_interval <- activity %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))

# Time series plot of the average steps per interval -----------------------
ggplot(avg_interval, aes(x = interval, y = avg_steps)) +
  geom_line(color = "#d7191c") +
  labs(title = "Average Daily Activity Pattern",
       x = "5-Minute Interval",
       y = "Average Steps")

# Identify the interval with the maximum average steps ---------------------
avg_interval %>% 
  filter(avg_steps == max(avg_steps))

```

## Imputing Missing Values
```{r impute-missing}
# Count missing values -----------------------------------------------------
sum(is.na(activity$steps))

# Compute mean steps for each interval ------------------------------------
interval_means <- activity %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))

# Fill in missing step values using interval means ------------------------
activity_imputed <- activity %>%
  left_join(interval_means, by = "interval") %>%
  mutate(steps = if_else(is.na(steps), mean_steps, steps)) %>%
  select(date, interval, steps)

# Verify that no NAs remain ------------------------------------------------
sum(is.na(activity_imputed$steps))
```
## Histogram and Statistics after imputing missing values
```{r histogram-after-imputation}
# Total steps per day after imputation ------------------------------------
daily_steps_imputed <- activity_imputed %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))

# Histogram of total daily steps after imputation -------------------------
ggplot(daily_steps_imputed, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "#1a9641", color = "white") +
  labs(title = "Total Steps per Day (After Imputation)",
       x = "Steps per Day",
       y = "Frequency")

# Mean and median after imputation ----------------------------------------
mean_imputed <- mean(daily_steps_imputed$total_steps)
median_imputed <- median(daily_steps_imputed$total_steps)

mean_imputed
median_imputed
```

## Are there differences in activity patterns between weekdays and weekends?
```{r weekday-weekend-patterns}
# Add a new factor variable for day type -----------------------------------
activity_imputed <- activity_imputed %>%
  mutate(day_type = if_else(wday(date) %in% c(1, 7), "weekend", "weekday"))

# Compute average steps by interval and day type ---------------------------
avg_by_daytype <- activity_imputed %>%
  group_by(day_type, interval) %>%
  summarise(avg_steps = mean(steps))

# Plot average steps for weekdays vs weekends ------------------------------
ggplot(avg_by_daytype, aes(x = interval, y = avg_steps, color = day_type)) +
  geom_line() +
  facet_wrap(~day_type, ncol = 1) +
  labs(title = "Average Steps per Interval: Weekday vs Weekend",
       x = "5-Minute Interval",
       y = "Average Steps") +
  theme_minimal() +
  theme(legend.position = "none")

```

## Thank You
